---
project:
title: "Getting familiar with the Family Resources Survey using R"
author: "UK Data Service"
date: last-modified
date-format: "MMMM YYYY"
mainfont: "Arial"
title-block: "plain"
title-block-banner: "white"
title-block-banner-color: "#742082"
format:
  html:
   toc: true
   smooth-scroll: true
   toc-location: left
css: frs.css
execute:
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

This exercise is part of the 'Getting started with ADD MODULE NAME' online module.

In the exercise, we examine data from the Family Resources Survey 2021/2022 financial year and explore:

-   Two of the four well-being variables available in the data: `HAPPYWB` and `ANXIOUS`.
    The other two variables available are LIFESAT and MEANING.

-   How do these two variables vary across age groups, and

-   How do these two variables vary across family type.

## Getting started

Data can be accessed from the [UK Data Service website](http://doi.org/10.5255/UKDA-SN-9073-1) following registration.

To follow along with the exercise you will need to download the SPSS data format from the UK Data Service website.
After saving the file on your machine, unzip everything and make sure the folder "UKDA-9073-spss" is placed in the folder which contains the R project FRS.Rproj and R script files FRS_get_familiar.qmd that you have obtained from the Github repository <!-- Add link to repository? -->

The FRS is a big survey, as a result, the data is organised in multiple files.
However, in this exercise, we will be using only two or three of these data files.

Using an R project (.Rproj) offers a structured, isolated, and reproducible working environment.

### Setting up R

To begin with, we need to load all R packages we will be using for this exercise.
When using R and writing R scripts, it is good practice to have all the packages needed for analysis at the beginning of an R script.

```{r}
library(dplyr) # Data manipulation functions
library(haven) # Importing Stata/SPSS files
library(Hmisc) # Extra statistical functions
library(tidyverse) #Data manipulation and visualization.
library(janitor) #Data cleaning and summary table
library(knitr) #Tables in Quarto
library(DescTools) #Weighting
#Load the data files we are using in this exercise
frs_adult2122 <- read_spss ("UKDA-9073-spss/spss/spss25/adult.sav")
frs_benunit2122 <- read_spss ("UKDA-9073-spss/spss/spss25/benunit.sav")

```

## Explore the data sets

To begin with, we will be focusing on exploring the "adult.sav" and "benunit.sav" datasets.
The "adult.sav" file contains 590 variables while the "benunit.sav" file has 275 variables.
We will only be using a few of these variables.
As a result, it is better for us to create two subsets of the data, containing only the variables we are interested in.

**NOTE:** These two files have different units of analysis.
The "adult.sav" file contains data at the individual level while the "benunit.sav" file has data at the benefit unit level (see the [FRS documentation](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=9073#!/documentation) for the definition of a benefit unit)

```{r}
#use a pipeline (the %>% operator) 
#and the "select" function from the dplyr package
#the code below only selects the variables we are interested in
frs_adult2122_short <- frs_adult2122 %>% select (SERNUM, BENUNIT, PERSON, GROSS4, HAPPYWB, ANXIOUS, SEX, IAGEGR4)

frs_benunit2122_short <- frs_benunit2122 %>% select (SERNUM, BENUNIT, FAMTYPBU)
```

Start by getting an overall feel of the two data sets.

We will start with the "adult" dataset.
Either inspect variables and cases in the data editor or use the code below to produce a summary of all the variables in the dataset.

```{r}

dim(frs_adult2122_short) #Gives the number of rows (observations) and columns (variables)

names(frs_adult2122_short) #List variable names in their actual order in the dataset

head(data.frame(frs_adult2122_short)) #Displays the first few rows of a data frame
#you can use tail(data.frame(frs_adult2122_short)) to display the last few rows of a data frame

```

Now let's also have a quick look at the "benunit" dataset.

```{r}

dim(frs_benunit2122_short) #Gives the number of rows (observations) and columns (variables)

names(frs_benunit2122_short) #List variable names in their actual order in the dataset

head(data.frame(frs_benunit2122_short)) #Displays the first few lines of a data frame

```

**Questions**

1.  What is the overall sample size of each of the data sets?

2.  How many variables are there in the short adult dataset?

Now, let's learn more about the available variables.

**Note**: In traditional statistical software packages like SPSS or Stata, categorical variables are often represented as numeric codes with attached labels that convey their substantive meaning.
R on the other hand allows for the use of alphanumeric variables or a specialised type of categorical variable known as 'factors'.
Factors in R can be ordered or unordered and are a fundamental aspect of R's categorical data handling.

The *Haven* R package, which we use in this course, facilitates the conversion of categorical data from SPSS or Stata to R.
While it can preserve the original numeric values it also contains attributes, which are special types of R objects with a name, accessible via the *attr()* function.
In this context, each variable has two key attributes: a 'label', which is a description of the variable, and 'labels', which are the value labels.

Furthermore, the *Haven* R package allows for the conversion of these imported numeric variables into R factors.
In this conversion, the factors are created with levels (i.e., categories) reflecting the value labels from SPSS or Stata.
This approach offers a streamlined method to handle and analyse categorical data in R, bridging the gap between different statistical software methodologies.

Let's examine the original variables' description and value labels.

```{r}

#the cat and names functions are only used in this case to make the output look better, but you do not need to use them
cat(attr(frs_adult2122_short$HAPPYWB,"label"))

names(attr(frs_adult2122_short$HAPPYWB,"labels"))

cat(attr(frs_adult2122_short$ANXIOUS,"label"))

names(attr(frs_adult2122_short$ANXIOUS,"labels"))

cat(attr(frs_adult2122_short$IAGEGR4,"label"))

names(attr(frs_adult2122_short$IAGEGR4,"labels"))

cat(attr(frs_benunit2122_short$FAMTYPBU,"label"))

names(attr(frs_benunit2122_short$FAMTYPBU,"labels"))


```

**Questions**

3\.
What do the variables measure and how?

## Merging data sets

Why do we merge data?
Merging datasets allows us to combine information from different sources to enhance the depth of our analysis.
This is particularly relevant when dealing with survey data that captures different dimensions of information at multiple levels, such as individual and benefit unit levels.

By merging data that captures information at different levels we can examine how individual characteristics interact with benefit-unit-level variables to influence outcomes of interest.
In this we might want to examine the relationship between individual happiness and family type.
This requires assigning the benefit unit level variable to every individuals within that unit, before we can analyse it further.

How do we merge data?
In order to effectively merge datasets you need to use common identifiers - unique keys present across all datasets involved in the merge.
In this case, the FRS documentation provides us with the necessary information: what are the available identifiers and how to use them depending which datasets we want to merge.

In our case the SERNUM and BENUNIT variables are needed to merge the two datasets.
Sernum represents the unique serial number of the household, and Benunit represents the identifier for the benefit unit (family) in the household.

```{r}
#specify which datasets to merge and which columns to use for merging
merged_data <- merge(frs_adult2122_short, frs_benunit2122_short, by = c("SERNUM", "BENUNIT"))
```

**Questions**

4\.
How many variables do we have in the new merge_data?

## Missing values

Let's now examine the frequency of our wellbeing variables.

We will temporarily convert the variables into factors using \`as_factor()\` and \`mutate()\` for a more meaningful output.

Review the frequency tables, examining the missing values.

```{r}
#the code below displays the results in a nice table  
merged_data %>%
  select(HAPPYWB, ANXIOUS) %>% # Select variables
  mutate(across(everything(), as.factor)) %>% # Convert all selected variables to factors
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Category") %>% # Pivot data for easier counting
  group_by(Variable, Category) %>%
  summarise(Count = n(), .groups = 'drop') %>% # Summarise counts
  pivot_wider(names_from = Variable, values_from = Count) %>% # Pivot data to have variables as columns
  adorn_totals("row") %>% # Optionally, add totals row
  kable() # Display as a nice table 
```

**Questions**

5\.
Why do you think there are so many system missing values (NA) for each of the well-being variables?

Note, you can use the documentation to check if needed (see \[[FRS documentation](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=9073#!/documentation)\]).

What does this mean when it comes to interpreting the percentages?

**Note:** For ease of interpretation, we are converting the original numeric `FAMTYPBU` variable into labeled factors using `as_factor()`, so that they directly display the value labels.

```{r}
merged_data$FAMTYPBU <- as_factor(merged_data$FAMTYPBU)
merged_data$IAGEGR4 <- as_factor(merged_data$IAGEGR4)

#drop the "Any other category" level from the variable as it did not apply to anyone in the data
merged_data$FAMTYPBU <- droplevels(merged_data$FAMTYPBU)

```

## Compare unweighted and weighted proportions

Let's examine the unweighted responses first.

The \`xtabs()\` function can be used for categorical variables and the \`summary()\` function for continuous ones.

```{r}
round(            ### Round the results to one decimal
  100*            ### Convert proportions to %  
    prop.table(   ### Compute proportions
      xtabs(~FAMTYPBU,merged_data) ### Compute frequencies
    ),
  1)

round(            ### Round the results to one decimal
  100*            ### Convert proportions to %  
    prop.table(   ### Compute proportions
      xtabs(~IAGEGR4,merged_data) ### Compute frequencies
    ),
  1)
#unweigthed summary results
summary(merged_data$HAPPYWB)
summary(merged_data$ANXIOUS)
```

What is the (unweighted) percentage of individuals living in an couple without children?
Would this be representative of the population?

Let's compare with the weighted frequencies.
We will use the `wtd.table()` from the `Hmisc` package.
The weights are specified after the variable for which we request the frequencies in the command below.

```{r}
# Raw output
wtd.table(merged_data$FAMTYPBU,weights=merged_data$GROSS4)

# Converted into proportions            
round(
  100*
    prop.table(
      wtd.table(merged_data$FAMTYPBU,weights=merged_data$GROSS4)$sum.of.weights),
  1)
```

Now what is the representative proportion of couples without children?
26.9, is this higher or lower than the unweighted proportion?

Let's look at the relation between family type and happiness and compare the unweighted and weighted results?

```{r}
#unweighted
merged_data %>% 
  group_by(FAMTYPBU) %>% 
  summarise (happiness = round(mean(HAPPYWB, na.rm=TRUE),2))

#weighted
merged_data %>% 
  group_by(FAMTYPBU) %>% 
  summarise (happiness = round(weighted.mean(HAPPYWB,GROSS4, na.rm=TRUE),2))
```

Is there a difference between the results?
Why do you thin there might or not be a difference?

```{r}
#unweighted
merged_data %>% 
  group_by(IAGEGR4) %>% 
  summarise (happiness = round(mean(HAPPYWB, na.rm=TRUE),2))

#weighted
merged_data %>% 
  group_by(IAGEGR4) %>% 
  summarise (happiness = round(weighted.mean(HAPPYWB,GROSS4, na.rm=TRUE),2))
```

## **Confidence intervals**

So far, we have just computed point estimates without worrying about their precision.
We can compute confidence intervals to indicate the precision (uncertainty) of our estimates.

**Confidence intervals for numerical variables:** Numerous R packages provide functionalities for calculating confidence intervals and standard errors for means.
However, here we emphasise manual computation to gain a deeper understanding of the underlying processes.

Under the assumption of simple random sampling, a 95% confidence interval for the mean is calculated as the mean plus or minus 1.96 times the standard error of the mean.
The standard error of the mean itself is the standard error of the mean (that is, the square root of its variance) divided by the square of the sample size.
Since we have functions for computing weighted means and variance in R, we can compute:

```{r}
m.p.happy<-wtd.mean(merged_data$HAPPYWB,weights=merged_data$GROSS4)
se.p.happy<-sqrt(wtd.var(merged_data$HAPPYWB,weights=merged_data$GROSS4))
n.happy<-sum(merged_data$GROSS4[!is.na(merged_data$HAPPYWB)])

ci.happy<-c(m.p.happy,m.p.happy-1.96*(se.p.happy/sqrt(n.happy)),
            m.p.happy+1.96*(se.p.happy/sqrt(n.happy)))
ci.happy

m.p.anxious<-wtd.mean(merged_data$ANXIOUS,weights=merged_data$GROSS4)
se.p.anxious<-sqrt(wtd.var(merged_data$ANXIOUS,weights=merged_data$GROSS4))
n.anxious<-sum(merged_data$GROSS4[!is.na(merged_data$ANXIOUS)])

ci.anxious<-c(m.p.anxious,m.p.anxious-1.96*(se.p.anxious/sqrt(n.anxious)),
              m.p.anxious+1.96*(se.p.anxious/sqrt(n.anxious)))
ci.anxious
```

**Questions**

6.  How happy are people overall on a scale of 0 to 10?

**Confidence intervals for categorical variables:** The `DescTools` package provides `MultinomCI()` a handy function to compute confidence intervals for multinomial proportions.

We need to provide `MultinomCI()` with two parameters: the frequencies for which we would like a confidence interval, and the confidence interval we want.

```{r}
# Calculate the frequency table of the variable
category_freq <- table(merged_data$FAMTYPBU)

# Calculate the multinomial confidence intervals
ci_results <-round(((MultinomCI(category_freq, conf.level = 0.95))*100),1)

# View results
ci_results
```

## Answers

1.  There are `r nrow(frs_adult2122_short)` cases in the adult dataset and `r nrow(frs_benunit2122_short)` in the benunit dataset.
2.  The total number of variables is `r ncol(frs_adult2122_short)` in the adult dataset and `r ncol(frs_benunit2122_short)` in the benunit dataset.
3.  For example, `HAPPYWB` records responses to the questions of "How happy did you feel yesterday?" on a scale of 'Not at all happy' to 'Completely happy'. The `FAMTYPBU` from the benunit dataset provides the family type.
4.  There are a total of `r ncol(merged_data)` variables in the new merged data.
5.  The wellbeing variables cannot be answered by proxy, so if the person is not present to answer then no answer is recorded.
6.  People are relatively happy, with an average of 7.4 out of 10.
